---
title: "Knitting Paper"
date: "2026-04-06"
description: "An R Script for Generating Knitting Patterns"
categories: [Viz, ggplot]
format: 
  html: 
    extends: _formats.yml
---

As a hobby knitter I have developed the script below as a tool to support the design of knitting patterns.
It can be especially helpful for shaping tasks such as raglan constructions or calculating stitch decreases for necklines. The script is based on your personal knitting gauge, you can adjust the measured values by modifying the variables `no_of_stitches_per_10_cm` and `no_of_rows_per_10_cm`.
At the end of the script, you will also find a few commented lines of code for printing the knitting paper at the correct scale. Please note that the resulting pdf might have very unusual sizes which requires additional adjustments.
Feel free to use, adapt, and improve the R script to suit your own projects.


```{r setup_and_load_packages, message = FALSE}
# R packages needed
packages <- c("ggplot2", "dplyr", "tibble", "grid", "gtable", "tidyr")

# Install packages not yet installed
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Loading packages
invisible(lapply(packages, library, character.only = TRUE))

```

```{r knitting_paper, include = TRUE, echo = TRUE, fig.width = 12}
#| fig-cap: Knitting paper based on gauge results

# gauge result
no_of_stitches_per_10_cm <- 26.5
no_of_rows_per_10_cm <- 37

# size of the output
size_x_cm <- 25
size_y_cm <- 17

output_cm <- tibble(x = c(0, size_x_cm), y= c(0, size_y_cm))

# size of a single stitch in x and y direction
delta_x <- 10/no_of_stitches_per_10_cm
delta_y <- 10/no_of_rows_per_10_cm

total_no_of_stitches <- round(size_x_cm / delta_x)
total_no_of_rows <- round(size_y_cm / delta_y)

#  grid defined by stitch size
lines_v <- tibble(x = (0 : total_no_of_stitches) * delta_x)
lines_h <- tibble(y = (0 : total_no_of_rows) * delta_y)

#  support grid for 10x10 cm
lines_v_dm <- tibble(x = seq(0, size_x_cm, by = 10))
lines_h_dm <- tibble(y = seq(0, size_y_cm, by = 10))

#  support grid for 10x10 stitches and lables
lines_h_10_stitches<- tibble(x = (1:floor(total_no_of_stitches / 10)) * 10 * delta_x)
lines_v_10_stitches  <- tibble(y = (1:floor(total_no_of_rows / 10)) * 10 * delta_y)

lines_v_10_stitches_labels <- lines_v_10_stitches %>%
  mutate(
    label = round(y/ delta_y) ) %>% 
  crossing(tibble(x = seq(15, total_no_of_stitches, by = 30) * delta_x))


# tibble for alternating rows 
alternating_rows <- tibble(
  ymin = (seq(0, total_no_of_rows, by = 2)) * delta_y,
  ymax = ymin + delta_y,
  xmin = 0,
  xmax = delta_x * total_no_of_stitches)


# knitting paper plot

output_cm %>%
  ggplot(aes(x = x, y = y)) +
  geom_blank() +
  
  geom_rect(
    data = alternating_rows,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    fill = "grey80",
    color = NA,
    inherit.aes = FALSE
  ) +
  
  geom_vline(data = lines_v, aes(xintercept = x), color = "grey60", linewidth = 0.4) +
  geom_hline(data = lines_h, aes(yintercept = y), color = "grey70", linewidth = 0.4) +
  
  geom_vline(data = lines_h_10_stitches, aes(xintercept = x), color = "grey60", linewidth = 0.6) +
  geom_hline(data = lines_v_10_stitches, aes(yintercept = y), color = "grey60", linewidth = 0.6) +
  
  geom_vline(data = lines_v_dm, aes(xintercept = x), color = "darkred", linewidth = 0.7) +
  geom_hline(data = lines_h_dm, aes(yintercept = y), color = "darkred", linewidth = 0.7) +

 
  geom_text(
    data = lines_v_10_stitches_labels,
    aes(x = x,  y = y, label = label),
    col = "grey40", 
    inherit.aes = FALSE,
    hjust = 0.5,
    vjust = 0.5,
    size = 3
  ) +

coord_fixed(xlim = c(0, size_x_cm), ylim = c(0, size_y_cm), ratio = 1) +
  
  
 scale_x_continuous(
  name = "Width (cm)",
  breaks = lines_v_dm$x,
  minor_breaks = seq(0, size_x_cm, by =1),
  labels = round(lines_v_dm$x),
  expand = expansion(add = 1*delta_x),
  sec.axis = sec_axis(
    ~ . / delta_x,
    name = "Stitch Number",
    breaks = seq(0, total_no_of_stitches, by = 10 * floor(floor(total_no_of_stitches/10)/4))
  )
) +

scale_y_continuous(
  name = "Length (cm)",
  breaks = lines_h_dm$y,
  labels = round(lines_h_dm$y),
  expand = expansion(add = 1*delta_y),
  sec.axis = sec_axis(
    ~ . / delta_y,
    name = "Row Number",
    breaks = seq(0, total_no_of_rows, by = 10 * floor(floor(total_no_of_rows/10)/4))
  ) 
) +
 

  labs( title  = "Knitting Paper", 
  subtitle = paste(no_of_stitches_per_10_cm, "Stitches per 10 cm/", no_of_rows_per_10_cm, "Rows per 10 cm"), 
  caption = "Enjoy Knitting!\n Origami Data Science Services\n") +
    
  theme(
        title =element_text(size = 18, face = 'bold'),
        
        axis.title.x = element_text(color = "darkred", size = 15 ),
        axis.title.y = element_text(color = "darkred", size = 15),   
        axis.title.x.top = element_text(color = "grey40", size = 15),    
        axis.title.y.right = element_text(color = "grey40", size = 15),  
        
        axis.text.x = element_text(color = "darkred", size = 12),
        axis.text.y = element_text(color = "darkred", size = 12), 
        axis.text.x.top = element_text(color = "grey40", size = 12), 
        axis.text.y.right = element_text(color = "grey40", size = 12), 
        
        
  ) -> plot_knitting_pattern

# Plot of knitting paper

plot_knitting_pattern

#
## Print knitting pattern to a pdf file
## Sizes of the pdf file correspond to a 1:1 scale with regard to size_x_cm and size_y_cm
#
# g <- ggplotGrob(plot_knitting_pattern)
## Identify panel 
# panel_index <- g$layout$name == "panel"
#
## fix the size of the plot
# g$widths[g$layout$l[panel_index]] <- unit(size_x_cm, "cm")
# g$heights[g$layout$t[panel_index]] <- unit(size_y_cm, "cm")
# total_width  <- sum(g$widths)
# total_height <- sum(g$heights)
#
## pfd with knitting pattern
## BE AWARE OF UNUSUAL PAPER SIZES!
#
## ggsave("knitting_paper_1_to_1.pdf",
#       plot = g,
#       width = convertWidth(total_width, "cm", valueOnly = TRUE),
#       height = convertHeight(total_height, "cm", valueOnly = TRUE),
#       units = "cm",
#       device = "pdf")

```

