---
title: "ECGs of Fibrillating Hearts"
date: "2026-03-22"
description: "Slow versus fast and regular versus irregular fibrillations"
categories: [ggplot, FibFreq, DataViz]
---

You are probably familiar with what an ECG looks like under a normal sinus rhythm. Ventricular fibrillation, however, is a very different signal: highly irregular, nonstationary, and difficult to characterize. Clinically, it is one of the most critical cardiac arrhythmias as it is fatal within minutes if untreated.


Recently, together with former colleagues, I published a [paper](https://doi.org/10.3389/frsip.2025.1707422) analysing ECG recordings from fibrillating murine hearts from a time-series perspective (see [{{< iconify simple-icons:github >}} GitHub](https://github.com/origami-dss)). I also developed the accompanying [{{< iconify simple-icons r >}} package ***FibFreq***](https://github.com/origami-dss/FibFreq) package, which implements methods for estimating the fibrillation frequency. The package further includes a dataset of 112 ECG recordings. Below, I visualise four of these signals. Each time series is one second in duration and consists of 1,000 data points.


```{r setup_and_load_data, message = FALSE}
if (!require(tidyverse)) {
  install.packages("tidyverse")
}
if (!require(FibFreq)) {
  library(devtools)
  install_github("origami-dss/FibFreq")
}
library(tidyverse)
library(FibFreq)
```

```{r plot_ECGs}
#| fig-align: center
#| fig-width: 8
#| fig-height: 6
# chose four ECGs for being plotted
ecgs_selected <- c(1, 37, 112, 13)

# tibble for representation of chosen ECGs
minitib <- function(no, i) {
  tibble(
    no = rep(no, 1000),
    time = 1:1000,
    ecg = MiceFibECGs[, i],
    orig_no = rep(i, 1000)
  )
}
data_fig_ecgs <- map2_dfr(1:length(ecgs_selected), ecgs_selected, minitib)

# tibble for labeling the ECGs
data_fig_ecgs_text <- data.frame(
  no = c(1, 2, 3, 4),
  label = paste("ECG No.", ecgs_selected)
)

data_fig_ecgs_y_scales <- data.frame(
  no = c(1, 1, 2, 2, 3, 3, 4, 4),
  time = rep(0, 8),
  ecg = c(-0.1, 0.2, -0.03, 0.03, -0.03, 0.03, -0.01, 0.02)
)

# plot the data
data_fig_ecgs |>
  ggplot(aes(x = time, y = ecg)) +
  geom_line(col = "darkblue", linewidth = 0.5) +
  geom_blank(data = data_fig_ecgs_y_scales) +
  facet_grid(rows = vars(no), scales = "free") +
  geom_text(
    data = data_fig_ecgs_text,
    mapping = aes(x = 0, y = Inf, label = label, hjust = 0, vjust = 1),
    fontface = "bold",
    size = 5
  ) +
  labs(
    x = "Time [ms]",
    title = "Four Examples of Murine ECGs (Source: R Package FibFreq)",
    caption = "Dr. Annette Witt (2026)\n Origami Data Science Services"
  ) +
  scale_y_continuous(NULL, expand = expansion(mult = c(0, 0)), breaks = NULL) +
  scale_x_continuous(
    limits = c(-5, 1020),
    expand = c(0, 0),
    breaks = (0:5) * 200
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 16),
    plot.title = element_text(hjust = 0.0),
    strip.text.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.y = element_blank(),
    axis.line.y = element_blank(),
  )
```

All four time series exhibit oscillatory behaviour, with an apparent increase in irregularity from top to bottom. The frequencies lie between 20 and 30 Hz. Intuitively, one would expect a more reliable estimate of the fundamental frequency for the topmost ECG, and a less certain estimate for the ECG shown in the bottom panel. The ***FibFreq*** package provides both time-domain and frequency-domain methods for estimating the fundamental frequency of a time series. In addition, it returns measures that characterise the irregularity of the oscillations. Consistent with this expectation, the frequency analysis yields more uncertain estimates for the time series at the bottom of the chart than for those at the top. Further details can be found in the associated [paper](https://doi.org/10.3389/frsip.2025.1707422) and in the documentation of the {{< iconify simple-icons r >}} package.