---
title: "Fundamental Frequency of the Sunspot Data"
date: "2026-03-25"
description: How regular is the 11 year sunspot cycle?
categories: [ggplot, FibFreq]
format: 
  html: 
    extends: _formats.yml
---

How to compute the frequency of an oscillatory time series? And what can you do if this oscillation is irregular?
In our recent paper (<https://doi.org/10.3389/frsip.2025.1707422>), we proposed and discussed four time domain techniques  and three frequency domain techniques for doing that.
They are provided in the accompanying R package ***FibFreq***. Below I will briefly explain two of these techniques.
Their functionality is illustrated for the solar cycle time series .
We begin with loading the sunspot data set and smoothing it:


```{r setup_and_load_data, message = FALSE}
if(!require(tidyverse)) install.packages("tidyverse")
if(!require(FibFreq)){
   library(devtools)
    install_github("origami-dss/FibFreq")}
library(tidyverse)
library(FibFreq)

# We load the monthly  sunspot data set
sunspot_data <- as.vector(sunspot.month)

# which has a sampling interval of 
delta_t <- 1./12

# As the time series is very noisy, we apply a 3-year moving average filter
filtered_sunspot_data <- stats::filter(sunspot_data, rep(1./36,36))

# This leads to leading and ending values NA. 
# As the FibFreq library does not handle them, they have to be removed.
smoothed_sunspot_data <- filtered_sunspot_data[!is.na(filtered_sunspot_data)]

```

We begin with identifying the local maxima of this time series and consider the average temporal distance between successive maxima as a the fundamental period.

```{r plot_sunspot_data_and_tehir_maxs, include = TRUE, echo = TRUE, fig.width = 6, fig.height = 4}

# The parameters of the maximum identification function have to be adjusted 
freq_M2M <- freq_max2max(smoothed_sunspot_data, delta_t = delta_t, span = 31)

time = (1:length(smoothed_sunspot_data)) * delta_t
timings_maxima_Max2Max = tibble(time = freq_M2M$maxima_sampled * delta_t, 
                                sunspots = smoothed_sunspot_data[freq_M2M$maxima_sampled])
tibble(time = time, sunspots = smoothed_sunspot_data) %>%
  ggplot(aes(x = time, y = sunspots)) +
  geom_line(col = "darkgreen") +
  geom_point(data = timings_maxima_Max2Max, aes(x = time, y = sunspots), col = "darkred") + 
  labs( x = "Time [month]", y = " ", 
        title = "Smoothed Sunspot Data with Identified Maxima", 
        caption = paste("Dr. Annette Witt\n Origami Data Science Services\n", format(Sys.Date(), format = "%d. %B %Y"))) +
  theme_minimal() +
  theme(strip.text.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(), 
        plot.margin = margin(t = 5,  # Top margin
                             r = 5,  # Right margin
                             b = 5,  # Bottom margin
                             l = 20))

```

The fundamental period based on max-to-max distances corresponds to  `1./freq_M2M$freq_max2max = `r format(1./freq_M2M$freq_max2max, digits = 3) ``
years which is very close to the expected 11-year cycle. 
The coefficient of variation of the intervals between successive maxima is
 `freq_M2M$c_v = `r format(freq_M2M$c_v, digits = 2) ``, or, equivalently  the standard deviation  of the max-to-max distances is ``r format(freq_M2M$c_v/freq_M2M$freq_max2max, digits = 2) `` years which indicates  a moderate level of irregularity of the oscillation.

The fundamental frequency can also be estimated by spectral techniques:

```{r spectral_techniques, fig.width = 6, fig.height = 4}
freq_MaxP <- freq_argmax_periodogram(smoothed_sunspot_data,
                                     delta_t = delta_t,
                                     f_min = 0.001,
                                     f_max = 0.3)
freq_Lorentz <- freq_Lorentz_fit(smoothed_sunspot_data,
                                 delta_t = delta_t,
                                 f_min = 0.001,
                                 f_max = 0.3)

tibble_Lorentz_plot <- as_tibble(freq_Lorentz$Lorentz_model)


tibble_Lorentz_plot %>%
  ggplot(aes(x = Frequency)) +
  geom_linerange( aes(ymin = 0, ymax = Fourier_Amps), col = "darkgreen", linewidth = 1.5) +
  geom_line( aes(y = modelled_Fourier_Amps), col = "darkorange", linewidth = 1, linetype = "dashed") +
  labs(x = "Frequency [1/yr]", 
       title = "Smoothed Sunspot Data: Fourier Amplitudes and Lorentz Model", 
       caption = paste("Dr. Annette Witt\n Origami Data Science Services\n", format(Sys.Date(), format = "%d. %B %Y"))) +
  theme_minimal() +
  theme(
    strip.text.y = element_blank(),
    axis.title.y=element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )
```
The largest Fourier amplitude is found at $f=`r format(freq_MaxP$freq_argmax_periodogram, digits = 2) `$  which corresponds to a period of $P = `r format(1./freq_MaxP$freq_argmax_periodogram, digits = 3) `$ yr. If a Lorentz model is fitted to the Fourier amplitudes, it is characterized by a Lorentz frequency of $f_{\text{Lorentz}} = `r format(freq_Lorentz$freq_Lorentz, digits = 2) `$ yr^-1^  ($P_{\text{Lorentz}} = 10.7$ yr) along with a Lorentz scale parameter of 
$\sigma_{\text{Lorentz}} =`r format(freq_Lorentz$scale_Lorentz, digits = 2) `$  yr^-1^
indicating that the underlying oscillation is moderately irregular. 

The ***FibFreq*** package contains more estimators for the fundamental frequency. Feel free to explore them ;-)
