---
title: "On the Frequency of the Sunspot Data"
date: "2026-03-25"
description: How regular is the 11 year sunspot cycle?
categories: [ggplot, FibFreq, Analysis]
---

How can we compute the frequency of an oscillatory time series, and what can we do when the oscillations are irregular? In our recent [paper](https://doi.org/10.3389/frsip.2025.1707422), we proposed and discussed four time-domain techniques and three frequency-domain techniques for addressing this problem. These methods are implemented in the accompanying ***FibFreq*** {{< iconify simple-icons r >}} package. Below, I briefly describe two of these approaches. Their use is illustrated with the sunspot cycle time series. We begin by loading and smoothing the sunspot dataset:


```{r setup_and_load_data, message = FALSE}
if (!require(tidyverse)) {
  install.packages("tidyverse")
}
if (!require(FibFreq)) {
  library(devtools)
  install_github("origami-dss/FibFreq")
}
library(tidyverse)
library(FibFreq)

# We load the monthly  sunspot data set
sunspot_data <- as.vector(sunspot.month)

# which has a sampling interval of
delta_t <- 1. / 12

# As the time series is very noisy, we apply a 3-year moving average filter
filtered_sunspot_data <- stats::filter(sunspot_data, rep(1. / 36, 36))

# This leads to leading and ending NA values.
# As the FibFreq library does not handle them, they have to be removed.
smoothed_sunspot_data <- filtered_sunspot_data[!is.na(filtered_sunspot_data)]
```

We begin with identifying the local maxima of this time series and consider the average temporal distance between successive maxima as a the fundamental period.

```{r plot_sunspot_data_and_their_maxs}
#| fig-align: center
#| fig-width: 8
#| fig-height: 6
#| 
# The parameters of the maximum identification function have to be adjusted
freq_M2M <- freq_max2max(smoothed_sunspot_data, delta_t = delta_t, span = 31)

time = (1:length(smoothed_sunspot_data)) * delta_t
timings_maxima_Max2Max = tibble(
  time = freq_M2M$maxima_sampled * delta_t,
  sunspots = smoothed_sunspot_data[freq_M2M$maxima_sampled]
)
tibble(time = time, sunspots = smoothed_sunspot_data) %>%
  ggplot(aes(x = time, y = sunspots)) +
  geom_line(col = "darkgreen") +
  geom_point(
    data = timings_maxima_Max2Max,
    aes(x = time, y = sunspots),
    col = "orange",
    alpha = 0.5,
    size = 4
  ) +
  geom_point(
    data = timings_maxima_Max2Max,
    aes(x = time, y = sunspots),
    col = "red"
  ) +
  labs(
    x = "Time [month]",
    y = " ",
    title = "Smoothed Sunspot Data with Identified Maxima",
    caption = "Dr. Annette Witt (2026)\n Origami Data Science Services"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 16),
    strip.text.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )
```

The fundamental period based on maximum-to-maximum distances is given by `1./freq_M2M$freq_max2max = `r format(1./freq_M2M$freq_max2max, digits = 3) `` years, which is very close to the expected 11-year cycle. The coefficient of variation of the intervals between successive maxima is `freq_M2M$c_v = `r format(freq_M2M$c_v, digits = 2) ``, or equivalently, the standard deviation of the maximum-to-maximum distances is ``r format(freq_M2M$c_v/freq_M2M$freq_max2max, digits = 2) `` years, indicating a moderate level of irregularity in the oscillation.

The fundamental frequency can also be estimated by spectral techniques:

```{r spectral_techniques}
#| fig-align: center
#| fig-width: 8
#| fig-height: 6
#| 
freq_MaxP <- freq_argmax_periodogram(
  smoothed_sunspot_data,
  delta_t = delta_t,
  f_min = 0.001,
  f_max = 0.3
)
freq_Lorentz <- freq_Lorentz_fit(
  smoothed_sunspot_data,
  delta_t = delta_t,
  f_min = 0.001,
  f_max = 0.3
)

tibble_Lorentz_plot <- as_tibble(freq_Lorentz$Lorentz_model)


tibble_Lorentz_plot %>%
  ggplot(aes(x = Frequency)) +
  geom_linerange(
    aes(ymin = 0, ymax = Fourier_Amps),
    col = "darkgreen",
    linewidth = 2
  ) +
  geom_line(
    aes(y = modelled_Fourier_Amps),
    col = "darkgoldenrod1",
    linewidth = 1,
  ) +
  labs(
    x = "Frequency [1/yr]",
    title = "Smoothed Sunspot Data: Fourier Amplitudes and Lorentz Model",
    caption = "Dr. Annette Witt (2026)\n Origami Data Science Services"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 16),
    strip.text.y = element_blank(),
    axis.title.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )
```
The largest Fourier amplitude occurs at $f=`r format(freq_MaxP$freq_argmax_periodogram, digits = 2) `$,  corresponding to a period of $P = `r format(1./freq_MaxP$freq_argmax_periodogram, digits = 3) `$ years. 
If a Lorentzian model is fitted to the Fourier amplitudes (orange line), it yields a Lorentz frequency of $f_{\text{Lorentz}} = `r format(freq_Lorentz$freq_Lorentz, digits = 2) `$ yr^-1^  ($P_{\text{Lorentz}} = 10.7$ yr) and a Lorentz scale parameter of 
$\sigma_{\text{Lorentz}} =`r format(freq_Lorentz$scale_Lorentz, digits = 2) `$  yr^-1^. The latter is approximately  12% of $f_{\text{Lorentz}}$, indicating moderately irregular underlying oscillation . 

The ***FibFreq*** package contains more techniques for estimating the fundamental frequency. Feel free to explore them ;-)
