---
title: "Event Series Analysis"
date: "2026-03-06"
description: "How to analyse and visualize rate changes in event series?"
categories: [ggplot, Analysis]
---

Event series are time series that record the number of events within equally sized time intervals. They are commonly used to model phenomena such as neuronal spike trains, hydrogeological extremes (e.g. earthquakes or floods), or the occurrence of ectopic heartbeats in an otherwise regular cardiac rhythm. Event series may exhibit either constant or time-dependent event rates. A key question is how to distinguish between these two cases.

To illustrate this, we consider two examples, beginning with an event series characterised by a constant rate. The dataset contains a substantial observational gap in the interval from 4000 to 5500.

```{r install_tidyverse, message = FALSE}
library(tidyverse)
# define the color scheme
cols <- c("Events" = "black", "Spline" = "red", "Constant" = "blue")
```

```{r synthetic_event_series_constant_rate}
#| fig-align: center
#| fig-width: 8
#| fig-height: 6
#| fig-cap: Estimated time-dependent event rate (red dashed line) with its 95% confidence band (grey) and the estimated constant rate (blue line); rug plot of events shown at the bottom.
#| 
set.seed(3)
events <- rpois(10000, 0.05)
event_series <- tibble(x = 1:length(events), y = events)
is.na(event_series$y) <- c(4000:5500)

events <- event_series |> filter(y > 0)
ggplot() +
  geom_rect(
    aes(xmin = 4000, xmax = 5500, ymin = -Inf, ymax = Inf),
    color = NA,
    fill = "olivedrab",
    alpha = 0.3
  ) +
  geom_rug(
    data = events,
    aes(x = x, col = "Events"),
    lwd = 0.05,
    show.legend = FALSE
  ) +
  geom_point(
    data = events[1, ],
    aes(x = x, y = y, col = "Events"),
    shape = 124,
    size = 5
  ) +
  geom_smooth(
    data = event_series,
    aes(x = x, y = y, col = "Spline"),
    formula = "y ~ x",
    method = "loess",
    lty = "dashed",
    na.rm = TRUE,
    alpha = 0.5,
    linewidth = 0.5
  ) +
  geom_smooth(
    data = event_series,
    method = "lm",
    na.rm = TRUE,
    formula = "y ~ 1",
    aes(x = x, y = y, col = "Constant"),
    se = FALSE,
    linewidth = 0.5
  ) +
  annotate(
    "text",
    label = "No Data",
    col = "white",
    x = 4500,
    y = 0.05,
    size = 8,
    angle = 90
  ) +
  coord_cartesian(
    ylim = c(0.0, 0.10),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    breaks = seq(0, 0.10, by = 0.02),
    minor_breaks = NULL
  ) +
  guides(
    col = guide_legend(
      override.aes = list(
        shape = c(NA, NA, 124),
        linetype = c("solid", "dashed", "blank"),
        linewidth = c(0.5, 0.5, 0),
        size = c(NA, NA, 4)
      )
    )
  ) +
  scale_color_manual(
    values = cols,
    breaks = c("Constant", "Spline", "Events")
  ) +
  labs(
    x = "Time",
    y = "Event Rate",
    title = "Event Rates for a Poisson Process\n",
    col = "Data/ Models",
    caption = paste(
      "Dr. Annette Witt\n Origami Data Science Services\n",
      format(Sys.Date(), format = "%d. %B %Y")
    )
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 16),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.margin = margin(t = 10, r = 5, b = 2, l = 5)
  )
```

The plot shows a rug plot of the event series at the bottom, along with the estimated time-dependent rate (red dashed line), its 95% confidence band (grey), and the estimated constant rate (blue line). The estimated rate is more uncertain at the beginning and the end of the event series, as well as in the vicinity of the observational gap; that is, the confidence band widens.

For event series with time-dependent (here, periodically varying) rates, however, the following features are likely to be observed:

```{r synthetic_event_series_oscillating_rate}
#| fig-align: center
#| fig-width: 8
#| fig-height: 6
#| fig-cap: The graph represents the same analysis as above, here for an event series with a non-constant event rate.
#| 
set.seed(1)
len <- 10000
rate <- tibble(rate = 0.05 + 0.025 * sin(0.0005 * 2 * pi * (1:len)))
event_series <- rate |>
  mutate(x = row_number(), y = map_dbl(rate, \(x) rpois(1, x)))
events <- event_series |> filter(y > 0)

ggplot() +
  geom_rug(
    data = events,
    aes(x = x, col = "Events"),
    lwd = 0.05,
    show.legend = FALSE
  ) +
  geom_point(
    data = events[1, ],
    aes(x = x, y = y, col = "Events"),
    shape = 124,
    size = 5
  ) +
  geom_smooth(
    data = event_series,
    aes(x = x, y = y, col = "Spline"),
    method = "loess",
    formula = "y ~ x",
    span = 0.1,
    lty = "dashed",
    alpha = 0.5,
    linewidth = 0.5
  ) +
  geom_smooth(
    data = event_series,
    method = "lm",
    formula = "y ~ 1",
    aes(x = x, y = y, col = "Constant"),
    se = FALSE,
    linewidth = 0.5
  ) +
  guides(
    col = guide_legend(
      override.aes = list(
        shape = c(NA, NA, 124),
        linetype = c("solid", "dashed", "blank"),
        linewidth = c(0.5, 0.5, 0),
        size = c(NA, NA, 4)
      )
    )
  ) +
  coord_cartesian(
    ylim = c(-0.00, 0.12),
    expand = FALSE,
  ) +
  scale_y_continuous(
    breaks = seq(0, 0.12, by = 0.02),
    minor_breaks = NULL
  ) +
  scale_color_manual(
    values = cols,
    breaks = c("Constant", "Spline", "Events")
  ) +
  labs(
    x = "Time",
    y = "Event Rate",
    title = "Event Rates for a Poisson Process\nwith a Time-Dependent Rate",
    col = "Data/ Models",
    caption = paste(
      "Dr. Annette Witt\n Origami Data Science Services\n",
      format(Sys.Date(), format = "%d. %B %Y")
    )
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 16),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_blank()
  )
```

Here, the constant rate (blue line) periodically lies outside the confidence band of the time-dependent rate (grey area).