---
title: "ECGs of Fibrillating Hearts"
date: "2026-03-22"
description: "Slow versus fast and regular versus irregular fibrillations"
categories: [Viz, ggplot, FibFreq]
---

You’re probably familiar with what an ECG looks like under normal (sinus) rhythm. Ventricular fibrillation, however, is a very different signal: highly irregular, nonstationary, and difficult to characterize. Clinically, it’s one of the most critical cardiac arrhythmias—fatal within minutes if untreated.

In our recent paper (<https://doi.org/10.3389/frsip.2025.1707422>), we analyze ECG recordings from fibrillating murine hearts from a time series perspective. I have written the accompanying R package ***FibFreq*** with methods for estimating the fibrillation frequency and which includes a data set of 112 ECG recordings. Below, I visualize a small selection of these signals.

Each time series is one second long, sampled at 1 kHz (1,000 data points). To reduce 50 Hz power line noise, we applied a fifth-order Butterworth low-pass filter with a 45 Hz cutoff.

```{r, message = FALSE}
if(!require(tidyverse)) install.packages("tidyverse")
if(!require(FibFreq)){
   library(devtools)
    install_github("origami-dss/FibFreq")}
library(tidyverse)
library(FibFreq)
```

```{r plot_ECGs, include = TRUE, echo = TRUE, fig.width = 6, fig.height = 4}
#| fig-cap: ECGs signals of murine fibrillating hearts.
ecgs_selected  <- c(1, 37, 112, 13)
minitib <- function(no, i) {
  tibble(no = rep(no, 1000), 
         time = 1:1000, 
         ecg = MiceFibECGs[,i], 
         orig_no = rep(i, 1000)) }

map2_dfr(1:length(ecgs_selected), ecgs_selected, minitib) -> data_fig_ecgs

data_fig_ecgs_text <- data.frame(
  no   = c(1,2,3,4),
  label = paste("ECG No.",ecgs_selected))

data_fig_ecgs_y_scales <- data.frame(
  no   = c(1,1,2,2,3,3,4,4),
  time = rep(0,8),
  ecg = c(-0.1, 0.2, -0.03, 0.03, -0.03, 0.03, -0.01, 0.02))

data_fig_ecgs %>%
  ggplot(aes(x=time, y=ecg)) +
  geom_line(col="darkblue", linewidth = 0.5) +
  geom_blank(data = data_fig_ecgs_y_scales) +
  facet_grid(rows = vars(no), scales = "free") +
  geom_text(data = data_fig_ecgs_text, 
            mapping = aes(x = 0, y = Inf, label = label, hjust = 0, vjust = 1 ), 
            fontface = "bold", size = 3) +
  labs( x = "Time [ms]", title = "Examples of Murine ECGs (Source: R Package FibFreq) \n",  
        caption = paste("Dr. Annette Witt\n Origami Data Science Services\n", format(Sys.Date(), format = "%d. %B %Y"))) +
  scale_y_continuous(NULL, expand = expansion(mult = c(0, 0)), breaks = NULL) +
  scale_x_continuous(limits= c(-01, 1020),  expand =  c(0, 0), breaks = (0:5)*200) +
  theme_minimal() +
  theme(strip.text.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        plot.margin = margin(t = 5,  # Top margin
                             r = 5,  # Right margin
                             b = 5,  # Bottom margin
                             l = 20)
  )
```

All four time series exhibit oscillatory behavior, with an apparent increase in irregularity from top to bottom. Intuitively, one would expect a more reliable estimate of the fundamental frequency for the topmost ECG, and a less certain estimate for the ECG shown in the bottom panel. The ***FibFreq*** package provides four time-domain and three frequency-domain techniques for estimating the fundamental frequency of a time series. In addition, it outputs quantities that characterize the irregularity of the oscillations .