---
title: "Correlations between Statistical Estimators"
date: "2026-03-28"
description: "How different are estimators of the same quantity?"
categories: [Viz, ggplot]
format: 
  html: 
    extends: _formats.yml
---
In data science we are often confronted  with the same quantity measured at different times or at different spatial locations, or with different devices or different statistical estimators the same quantity. How to dependent are the results? In this post, I will vizualize it.

As test data I will use a blood presure data compiled in Ghana, were the blood pressure of a large cohort was recorded for most patients three times. Each measurement consist of two numbers: the systolic and the diastoloc pressure.

```{r setup_and_load_packages, message = FALSE}
# Packages required for this blog entry
packages <- c("tidyverse", "bp", "scico", "corrViz", "GGally")

# Install packages not yet installed
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) install.packages(packages[!installed_packages])

# Loading packages
invisible(lapply(packages, library, character.only = TRUE))
```


First, the blood pressure data are loaded and preprocessed:

```{r load_and_preprocess_data}
data(bp_ghana)

bp_ghana %>% 
  filter(!is.na(SBP) & !is.na(DBP)) %>% # remove all rows with NA value od the blood pressure
  group_by(ID) %>% 
  mutate(n= n()) %>% 
  filter(n==3) %>%  # remove all patients with incomplete bp measurements
  ungroup()   ->  bp_data  
# show the head of the table
head(bp_data)

my_colors <- c("Systolic BP" = "darkblue", "Diastolic BP" = "coral")
```

This results in a table with 547 different patient IDs. 

In a first graphical presentation I  present  blood pressure data as boxplots in dependence on the elapsed time.



```{r plot_boxplots, include = TRUE, echo = TRUE, fig.width = 6, fig.height = 4}

COLS = scico(6, palette = 'oleron')
# prepare data table for boxplot
bp_data  %>% 
  select(ID, Time_Elapsed, SBP, DBP) %>%
  mutate(time = case_when(Time_Elapsed == "Baseline" ~ 1,
                         Time_Elapsed == "6Months" ~ 2, Time_Elapsed == "12Months" ~ 3)) %>%
  pivot_longer(names_to = "SBP_or_DBP", cols = c(SBP, DBP)) %>%
    mutate(SBP_or_DBP = factor(SBP_or_DBP, levels = c("SBP", "DBP"), labels = c("Systolic BP","Diastolic BP"))) -> bp_data_for_boxplot

bp_data_for_boxplot %>%
  ggplot(aes(x = as.factor(time), y = value, fill = SBP_or_DBP)) +
   geom_point(position = position_jitterdodge(dodge.width = 0.6, jitter.width = 0.5),  
   alpha = 0.9, size = 0.3, col = "grey") +
  geom_boxplot(alpha = 0.7, width = 0.6, outliers = FALSE, size = 0.3) +
  stat_summary(position = position_dodge(width = 0.6), 
  fun = mean, geom='point', shape = 18, size = 3) + 
  scale_fill_manual(values = my_colors) +
  labs(title = "Blood Pressure (mmHg) over Time", 
  subtitle = "Ghana Blood Pressure Data Set Provided by the bp R-Package", 
  y = NULL, 
  x = "Time Elapsed",
  caption = paste("Dr. Annette Witt\n Origami Data Science Services\n", format(Sys.Date(), format = "%d. %B %Y")),  
  fill = NULL) +
 scale_x_discrete(labels = c("Baseline", "6 Months" , "12 Months")) + 
  theme_minimal() +
  theme(
    panel.grid.minor.x = element_blank(), 
    panel.grid.major.x = element_blank()
  )

```

This boxplot visualization shows the distribution of both systolic and diastolic blood pressure measurements across the three time points. The jittered points show individual measurements, while the diamonds indicate the mean values for each group.

```{r}
bp_data %>% 
  select(ID, Time_Elapsed, SBP, DBP) %>% 
  mutate(Time_Elapsed = factor(Time_Elapsed)) %>%
  pivot_longer(cols = c("SBP", "DBP"),  names_to = "SBP_or_DBP", values_to = "BP") %>%
  pivot_wider(names_from = Time_Elapsed, values_from = "BP") %>%
  mutate( SBP_or_DBP = factor(SBP_or_DBP, levels = c("SBP", "DBP"), labels = c("Systolic BP","Diastolic BP"))) -> bp_for_corrplot

odss_points_range <- function(data, mapping) { 
  ggplot(data = data, mapping = mapping) + 
    geom_abline(color = "darkgrey")+ 
    geom_point(aes(color = SBP_or_DBP), size = 0.5, alpha = 0.5) +
    scale_y_continuous(limits = c(45, 200)) +
    scale_x_continuous(limits = c(45, 200)) +
    scale_color_manual(values = my_colors)
}

odss_dens_range <- function(data, mapping) { 
  ggplot(data = data, mapping = mapping) +  
    geom_density(aes(fill = SBP_or_DBP), alpha = 0.5, linewidth = 0.5) +
    scale_fill_manual(values = my_colors)
}

odss_hist_range <- function(data, mapping) { 
  ggplot(data = data, mapping = mapping) +  
    geom_histogram(aes(fill = SBP_or_DBP, color = SBP_or_DBP), position = "identity", binwidth = 10, alpha = 0.2, linewidth = 0.25) +
    scale_fill_manual(values = my_colors) +
    scale_color_manual(values = my_colors)
}

odss_cor <- function(data, mapping) {
  x_var <- rlang::as_name(mapping$x)
  y_var <- rlang::as_name(mapping$y)
  
  # Skip correlation calculation for diagonal elements (same variable)
  if (x_var == y_var) {
    return(ggplot() + theme_void())
  }
  
  cor_sbp <- cor(data[data$SBP_or_DBP == "Systolic BP", x_var], 
                 data[data$SBP_or_DBP == "Systolic BP", y_var], 
                 use = "complete.obs")
  cor_dbp <- cor(data[data$SBP_or_DBP == "Diastolic BP", x_var], 
                 data[data$SBP_or_DBP == "Diastolic BP", y_var], 
                 use = "complete.obs")
  
  ggplot() + 
    # Background rectangles for better visibility
#    annotate("rect", xmin = 100, xmax = 100, ymin = 0.55, ymax = 0.75, 
#             fill = "white", alpha = 0.8, color = "navy", linewidth = 0.5) +
#    annotate("rect", xmin = 0.1, xmax = 0.9, ymin = 0.25, ymax = 0.45, 
#             fill = "white", alpha = 0.8, color = "orange3", linewidth = 0.5) +
    # Text annotations with larger size
    annotate("text", x = 100, y = 80, 
             label = paste("SBP:", round(cor_sbp, 2)), 
             color = "navy", size = 4) +
    annotate("text", x = 100, y = 120, 
             label = paste("DBP:", round(cor_dbp, 2)), 
             color = "orange3", size = 4) +
    theme_void()
}

# Remove the scales from the main ggpairs call
bp_for_corrplot %>%
    ggpairs(columns = c('Baseline', '6Months', '12Months'), 
    lower = list(continuous = odss_points_range),
    diag = list(continuous = odss_hist_range),
    upper = list(continuous = odss_cor),
    columnLabels = c('Baseline', '6 Months', '12 Months')) +
  ggplot2::coord_cartesian(xlim = c(45, 200)) +
  theme_minimal() +
  theme(
  panel.grid.major.y = element_blank(), 
  panel.grid.major.x = element_blank(),   
  panel.grid.minor.x = element_blank(),   
  panel.grid.minor.y = element_blank()   
  )
```